---
title: "S3 Methods in R - Part 1"
author: "Rossana Tatulli"
date: "`r Sys.Date()`"
format: 
  revealjs:
    slide-number: true
    theme: white  # Try 'moon' for dark mode, 'black' for high contrast
    transition: fade
    transition-speed: fast
    center: true  # Center align content
    highlight-style: github  # Better syntax highlighting
    chalkboard: true  # Enable drawing on slides
    controls: true  # Show navigation controls
    progress: true  # Show progress bar
    history: true  # Allow navigating back/forward
    width: 1200  # Increase slide width
    height: 800   # Increase slide height
    font: "Fira Code"  # Better font for code readability
    css: "custom.css"  # External CSS for further styling
editor: visual
execute:
  echo: true
  error: TRUE
  warning: FALSE  # Hide warnings for cleaner slides
---

## Intro on OOP and S3 methods

-   **OOP**: based on objects, which have a class, - *what the object is* -, and a method - *what the object does*
-   classes are organized into **hierarchies** so that if a method is missing for a certain class, it is inherited from the parent
-   Main goal of OOP: **polymorphism**, so that you can consider the function interface separately from the function implementation
-   S3 methods are most commonly used OOP system in CRAN packages, they are an implementation of ***functional OOP***:
    -   methods belong to generic functions
    -   method calls look like ordinary functions calls

## The basics

-   An S3 object is a base type with at least a `class` attribute(e.g., factor)

-   It behaves differently from its underlying base type whenever it’s passed to a **generic function**, which**:**

    -   uses a different implementation (the **method**) depending on the class of an argument

    -   finds that method by performing **method dispatch**

You can use [`sloop::s3_dispatch()`](https://sloop.r-lib.org/reference/s3_dispatch.html) to see the process of method dispatch:

```{r echo = TRUE}
library(sloop)
f <- factor(c("a", "b", "c"))
s3_dispatch(print(f))
```

## Classes

-   no formal definition of a class, to make an object an instance of a class, you simply set the **class attribute**

```{r}
# Create and assign class in one step
x <- structure(list(), class = "my_class")

# Create, then set class
x <- list()
class(x) <- "my_class"
```

-   You can determine the class of an S3 object with [`class(x)`](https://rdrr.io/r/base/class.html), and see if an object is an instance of a class using [`inherits(x, "classname")`](https://rdrr.io/r/base/class.html).

```{r}
class(x)
inherits(x, "my_class")

inherits(x, "your_class")
```

## **Class name guidelines**: {.smaller}

-   **Any string can be a class name**, but it’s advised to keep it simple by using only **letters** and **underscores** (`_`). Avoid using periods (`.`) in the class name because it could be confused with the period separator used between a generic name and a class name in R (e.g., `generic.class`).

-   When using classes within a package, it’s a good idea to **include the package name in the class name** (e.g., `mypackage_myclass`). This reduces the risk of class name collisions

## **Three recommended functions for class management**: {.smaller}

1\. **Constructor: `new_myclass()`**

-   **Purpose**: The constructor is a **low-level function** that is responsible for **creating new objects** of your class with the correct internal structure.

2\. **Validator: `validate_myclass()`**

-   **Purpose**: The validator is used to **check** if the object’s internal values are valid or meet certain conditions.

-   **What it does**: It performs more **computationally expensive checks** to ensure that the object's fields contain **correct data**

3\. **Helper: `myclass()`**

-   **Purpose**: The helper function is a **convenience function** for users to easily create objects of your class. It is typically a wrapper around the constructor function.

## Example

```{r}
# Constructor
new_my_factor <- function(values, levels) {
  stopifnot(is.character(values), 
            is.character(levels))
  structure(
    list(values = values, 
         levels = levels),
    class = "my_factor"
  )
}

# Validator
validate_my_factor <- function(x) {
  if (!all(x$values %in% x$levels)) {
    stop("All values must be in levels.")
  }
  x
}

# Helper function
as_my_factor <- function(x, levels = unique(x)) {
  x <- as.character(x)
  validate_my_factor(new_my_factor(x, levels))
}

# Example usage
f <- as_my_factor(c("low", "medium", "high"), levels = c("low", "medium", "high"))
print(f)

```

## Generics and methods - method dispatch {.smaller}

-   every generic calls `UseMethod` to perform method dispatch. `UseMethod()` takes two arguments: the name of the generic function (required), and the argument to use for method dispatch (optional)
-   it creates a vector of method names, `paste0("generic", ".", c(class(x), "default"))`, and then looks for each potential method in turn..see this in action with `sloop::s3_dispatch()`
-   “default” class is not a real class, but is included to define a standard fallback that is found whenever a class-specific method is not available.

Most generics are very simple, and consist of only a call to `UseMethod()`, for example see

```{r}
mean
```

## Finding methods {.smaller}

To find all methods defined for a generic or associated with a class, use `sloop::s3_methods_generic()` and `sloop::s3_methods_class()`

```{r}

s3_methods_generic("mean")

s3_methods_class("ordered")

```

## Creating methods {.smaller}

-   **Best Practice**: Only create methods for generics or classes that you own (or collaborate with the author of the package/class to add your method).
-   **Argument Matching:** Your method should have the same argument signature as the generic, except when using ... for flexibility.
-   **Ellipsis (...)**: Using ... in methods allows you to accept additional arguments, but be careful because misspelled arguments will be silently ignored, which could cause hidden issues.

--\> we will see more of this in the practical session next time

------------------------------------------------------------------------

## Inheritance {.smaller}

Mechanism through which **different classes** can **share behavior**, powered by 3 ideas:

1\. Class can be a character vector

2\. If method not found for the first class, R looks to method for the second class

3\. A method can delegate work by calling `NextMethod()`, "-\>"

------------------------------------------------------------------------

```{r}

# Example with ordered 
# ordered() creates an ordered factor with a given input 

# 1. ----- #
ordered(1:4)
class(ordered(1:4))

# (Even though both objects contain the same values, ordered() ensures that comparisons (<, >) are meaningful:)

# 2. ----- #
x <- 1:4
s3_dispatch(print(ordered("x")))

# 3. ----- #
s3_dispatch(ordered("x")[1])

```

------------------------------------------------------------------------

***Output takeaways***

[*Step 1: Printing*]{.underline}

\- The print() function is called on the result of ordered("x")\[1\].

\- R first tries print.factor, which is the specialized print method for factors (including ordered factors).

\- The =\> symbol means this method was used.

\- The \* before `print.default` means it was skipped, because `print.factor` was found.

--\> Since an ordered factor is also a factor, R uses print.factor to display it

------------------------------------------------------------------------

[*Step 2: Subsetting*]{.underline}

\- When `ordered("x")[1]` is evaluated, R looks for a `[.ordered` method (a special method for subsetting ordered factors).

\- No =\> appears next to `[.ordered`, meaning it was not found.

\- R then falls back to `[.factor`, which exists and is used (=\> means this method was applied)

\- `-> [ (internal)` means that the base "'\\\\['" operator was used at the lowest level

--\> Ordered factors don’t have a specialized `[.ordered` method, so they rely on `[.factor`.

--\> The actual subsetting operation is ultimately handled by the internal `[` function.

------------------------------------------------------------------------

::: {.callout-caution collapse="true"}
## Why delegation to `-> [ (internal)`

In R’s S3 method dispatch system:

-   R first looks for `[.ordered`, but it doesn't exist.
-   It then finds `[.factor` and uses it (=\> means it was applied).
-   Inside `[.factor`, R calls `NextMethod()`, which means:
    -   It delegates the call to `[.default` (if it exists) or the base `[` function.
    -   Since `[.default` is not used here, it goes directly to `[ (internal)`.

--\> 💡 Key takeaway:

`-> [ (internal)` means that `[.factor` did not completely handle the operation itself but instead delegated the actual work to the base R \``[` function (written in C). This delegation is likely happening via `NextMethod()`
:::

## Relationship between the classes that appear together in a class vector {.smaller}

In our example, `ordered` is a **subclass** of `factor` and `factor`is a **superclass** of `ordered`.

It is recommend that you adhere to two principles when creating a subclass:

1.  Sub-class and class belong to the **same base type**
2.  **Attributes** of the subclass should be a **superset** of the attributes of the **superclass**.

------------------------------------------------------------------------

**Example**

```{r}
# Function to create a "labelled_factor" subclass
labelled_factor <- function(x, label) {
  # Ensure the base type remains the same as the superclass (factor)
  if (!inherits(x, "factor")) {
    stop("x must be a factor")
  }
  
  # Add an extra attribute ("label") while preserving existing attributes
  structure(x, class = c("labelled_factor", "factor"), label = label)
}

# Example usage
x  <- factor(c("low", "medium", "high"))
lf <- labelled_factor(x, label = "Risk Level")

# Check structure
str(lf)
```

## `NextMethod()` {.smaller}

Let's start with an example

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

------------------------------------------------------------------------

However, the default `[` method does not preserve the class

```{r}
s3_dispatch(x[1])
x[1]

```

------------------------------------------------------------------------

Need to define a `[.secret` method:

Option 1:

```{r}
`[.secret` <- function(x, i) {
  new_secret(x[i])
}
```

Option 2:

```{r}
`[.secret` <- function(x, i) {
  x <- unclass(x)
  new_secret(x[i])
}
x[1]
```

::: callout-note
-   Can you guess why [Option 1]{.underline} does not work?

-   Why is [Option 2]{.underline} not efficient? (hint: *copy-on-modify* mechanism)
:::

------------------------------------------------------------------------

With `NextMethod()` instead

```{r}
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}
x[1]
s3_dispatch(x[1])
```

-   **`NextMethod()` tracks the list of potential next methods:** this means that R keeps track of the method dispatch chain, which consists of all the possible methods that could be invoked next for the object being processed

-   When a method is called, R checks the class of the object and finds the corresponding method. If the method defines a call to `NextMethod()`, R will **continue searching for the next method in the chain**, considering the inheritance hierarchy and class structure

## Allowing sub-classes {.smaller}

**Subclassing** might allows for more flexible and extensible code. To allow subclasses, the parent constructor needs to have`...`and `class` arguments:

```{r}
new_secret <- function(x, ..., class = character()) {
  stopifnot(is.double(x))

  structure(
    x,
    ...,
    class = c(class, "secret")
  )
}
#note that the class argument is used to attach the class to the object, and "secret" is added to the class chain.
```

------------------------------------------------------------------------

-   Subclasses (like `supersecret`) can be created by simply calling the parent constructor with the `class` argument modified to the subclass name.

```{r}
new_supersecret <- function(x) {
  new_secret(x, class = "supersecret")
}

```

------------------------------------------------------------------------

To allow inheritance, you need to think about your methods, as you can no longer use the constructor. See the example below:

```{r}
x2 <- new_supersecret(c(15, 1, 456))
x2

`[.secret` <- function(x, ...) {
  new_secret(NextMethod())
}

x2[1:3]
```

We need to refine the `[.secret` method, because currently it always return a `secret()` even when given a supersecret

-   need to use the vctrs package –\> we will see thisin the next session

-   general idea: using `vctrs::vec_restore()` helps manage subclassing: when an object loses its subclass information (e.g., after subsetting), it can be restored to the correct class.

## More on `UseMethod`, `NextMethod`, and `default` method in S3 {.smaller}

`UseMethod()`:

-   the mechanism that triggers method dispatch in S3 -dispatch appropriate function based on the class of the object passed to it
-   It's used in the **generic function**

Consider this example, what rules of function evaluation does `UseMethod()` violate?

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}

g.default <- function(x) c(x = x, y = y)

x <- 1
y <- 1

g(x)
g.default(x)
```

## Comments {.smaller}

-   x remains 1 because arguments are passed unchanged to methods.
-   However, local variables in `g()` (excluding function arguments) are accessible inside the method: y is taken from the local environment of `g()`, where it was set to 10, not from the global environment.
-   When `UseMethod()` dispatches a method, it preserves the environment where the generic function was called.
    -   Arguments (x) are explicitly passed, so they cannot be modified by the generic
    -   Other variables (y) inside the generic function are still available in the method's evaluation environment

2 takeaways: 1. Arguments are normally evaluated in the function’s environment, but **`UseMethod()` ignores changes inside the generic** 2. Normally, **local variables** in a function are not visible in **called functions**, but `UseMethod()` makes them available ---

`NextMethod()`:

-   used inside a method to call the next method in the dispatch chain
-   If a method is called for a subclass, the `NextMethod()` function is used to call the corresponding method of the superclass, keeping the method dispatch chain intact

------------------------------------------------------------------------

`default` methods:

-   Fallback methods that are used when no more specific method is found for a particular class
-   **useful** when you want to provide a generic implementation of a function for all object types, while still allowing specific classes to override the method with specialized behavior
-   **necessary** if you want to ensure that a function works for any object, even if that object doesn’t have a specific method defined for it
-   **unnecessary** when you want to ensure specific behavior for each subclass or class in a way that does not fall back on a generic implementation

------------------------------------------------------------------------

Example:

```{r define, error=TRUE}
# Define generic class
area <- function(x) {
  UseMethod("area", x)
}

# Define the specific methods
# Circle area
area.circle <- function(x) {
  pi * x$radius^2
}
# Square area
area.square <- function(x) {
  x$side^2
}
# Rectangle area
area.rectangle <- function(x) {
  x$length * x$width
}

# Define the Default Method
area.default <- function(x) {
  stop("No method for calculating area for this class.")
}
```

------------------------------------------------------------------------

```{r call-objs, error=TRUE}
# Create some objects and call the function

# Create a circle
circle <- list(radius = 5)
class(circle) <- "circle"

# Create a square
square <- list(side = 4)
class(square) <- "square"

# Create a rectangle
rectangle <- list(length = 6, width = 3)
class(rectangle) <- "rectangle"

# Create a generic object (no specific class)
unknown_shape <- list()
class(unknown_shape) <- "unknown"

# ________ ####
# Calling area for specific shapes
cat("Area of circle:", area(circle), "\n")        # Should calculate area of the circle
cat("Area of square:", area(square), "\n")        # Should calculate area of the square
cat("Area of rectangle:", area(rectangle), "\n")  # Should calculate area of the rectangle

# Calling area for an object with no specific class
cat("Area of unknown shape:", area(unknown_shape), "\n")  # Should trigger default method

```

## S3 methods - application with {vctrs} - intro

**Overview**

-   Main goals of {vctrs}

-   Creating S3 classes with {vctrs}

-   Understanding and Implementing Inheritance

-   Method Dispatch and Generic Functions

## {vctrs} {.smaller}

It provides tools to ensure that functions behave consistently with respect to inputs of varying length and type.

**Goals of vctrs**:

1\. Type stability

2\. Size stability

3\. Make it easier to build new S3 classes: {vctrs} gives a clear development path for creating an S3 class

```{r}
library(vctrs)
library(rlang)
```

```{r, error=TRUE}
# vec_c() behaves similarly to c() for atomic vectors 
vec_c(TRUE, 1)
vec_c(1L, 1.5)

# but is stricter when converting to character 
vec_c(1.5, "x")
c(1.5, "x")

```

these vctrs functions allow you to declare the desired "prototype" of the output

```{r, error=TRUE}
# for example, make the output look like an empty character vector
#vec_c(1.5, "x", .ptype = character())
```

Instead of implicitly coercing to a type, we cast to a specific type

## Example with factors

```{r, error=TRUE}
fa <- factor("a")
fb <- factor("b")
vec_c(fa, fb)
vec_c(fb, fa) # different levels 

# if you try to combine a factor and an ordered factor you'll get an error
oa <- ordered("a")
vec_c(fa, oa)
# explicitly declaring the type of output
vec_c(fa, oa, .ptype = character())
```

## 1. TYPE STABILITY {.smaller}

1.  The output type should only depend on the types of the input (in base R, `ifelse` violates this)

```{r}
# Output changes depending on type of logical element 
typeof(ifelse(NA, 1, "a"))
typeof(ifelse(NA, 1, "a"))
typeof(ifelse(NA, 1, "a"))
# undesirable bevause you don't know what type of output to expect
```

2.  The order of the inputs in vectors or functions should not affect the output type

```{r}
today <- as.Date("2025-03-24")
c(NULL, today)
c(today, NULL)
```

3.  It uses consistent rules

```{r}
c(fa, fb)
unlist(unlist(fa, fb))

# hard to have a mental model to guess what the operation is going to return
```

## 2. SIZE STABILITY

Size-stability = consistent recycling rules. Given the sized of the inputs you should be able to predict the sizes of the output

For example, when you combine two vectors of different sizes.

```{r}
c(1, 2, 3, 4) + c(10, 20)
# Shorter vector (c(10, 20)) is repeated to match length 4
c(1, 2, 3, 4, 5) + c(10, 20)
# Warning: longer object length is not a multiple of shorter object length


# TO COMPLETE - SEE VIGNETTE
```

## 3. EASIER WAYS OF CREATING NEW VECTORS CLASSES

For example, in R you might need to define methods for a lot of different generics. {vctrs}, instead, gives some documentation/support: e.g., which methods you need to define depending on what you want the vectors to do.

**Basics:** create a new vector class by calling `new_vctr()` , this creates an object of class `vctrs_vct` which has a number of methods

## Percent class

Let's make a make a `percent` class, i.e., a double vector that is printed as a percentage. Start by defining a low-level [constructor](https://adv-r.hadley.nz/s3.html#s3-constrcutor) to check types and/or sizes and call [`new_vctr()`](https://vctrs.r-lib.org/reference/new_vctr.html). `percent` is built on a double vector of any length and doesn’t have any attributes.

```{r}
new_percent <- function(x = double()) {
  if (!is_double(x)) {
    abort("`x` must be a double vector.")
  }
  new_vctr(x, class = "vctrs_percent")
}

x <- new_percent(c(seq(0, 1, length.out = 4), NA))
x

str(x)
```

Then define the helper

```{r}
percent <- function(x = double()) {
  x <- vec_cast(x, double())
  new_percent(x)
}
```

For the convenience of your users, consider implementing an `is_percent()` function:

```{r}
is_percent <- function(x) {
  inherits(x, "vctrs_percent")
}
```

## `format` method

first method for every class should almost always be a [`format()`](https://rdrr.io/r/base/format.html) method. This should return a character vector the same length as `x` The easiest way to do this is to rely on one of R’s low-level formatting functions like [`formatC()`](https://rdrr.io/r/base/formatc.html):

```{r}
format.vctrs_percent <- function(x, ...) {
  out <- formatC(signif(vec_data(x) * 100, 3))
  out[is.na(x)] <- NA
  out[!is.na(x)] <- paste0(out[!is.na(x)], "%")
  out
}

x
```

(Note the use of [`vec_data()`](https://vctrs.r-lib.org/reference/vec_data.html) so [`format()`](https://rdrr.io/r/base/format.html) doesn’t get stuck in an infinite loop)

## \>

For optimal display, recommended to define an abbreviated type name. This is used in tibbles and in [`str()`](https://rdrr.io/r/utils/str.html):

```{r}
vec_ptype_abbr.vctrs_percent <- function(x, ...) {
  "prcnt"
}

tibble::tibble(x)

str(x)
```

## Casting and coercion

The next set of methods you are likely to need are those related to coercion and casting.

-   When the change happens *implicitly* (e.g in [`c()`](https://rdrr.io/r/base/c.html)) we call it **coercion**; when the change happens *explicitly* (e.g. with `as.integer(x)`), we call it **casting**.

-   One of the main goals of vctrs: to put coercion and casting on a robust theoretical footing so it’s possible to make accurate predictions about what (e.g.) `c(x, y)` should do when `x` and `y` have different prototypes.

```{r}

# vec_ptype2(x, y)
# defines possible set of coercions. It returns a prototype if x and y can be safely coerced to the same prototype; otherwise it returns an error.

vec_ptype2(percent(), NA)
vec_ptype2(NA, percent())


# vec_cast(x, to) defines the possible sets of casts. It returns x translated to have prototype to
```

## Coercion for the case of identical class

It is always a good idea to provide an explicit coercion method for the case of identical classes. So we’ll start by saying that a `vctrs_percent` combined with a `vctrs_percent` yields a `vctrs_percent`

```{r}
vec_ptype2.vctrs_percent.vctrs_percent <- function(x, y, ...) new_percent()
```

## Other methods

define methods that say that combining a percent and double should yield a double.

```{r}
vec_ptype2.vctrs_percent.double <- function(x, y, ...) double()
vec_ptype2.double.vctrs_percent <- function(x, y, ...) double()
# double dispatch to check

# check what we've implemented
vec_ptype_show(percent(), double(), percent())
```

## Actual conversion: `vec_cast` {.smaller}

```{r}
# provide a method to cast a percent to a percent
vec_cast.vctrs_percent.vctrs_percent <- function(x, to, ...) x

# And then for converting back and forth between doubles.
vec_cast.vctrs_percent.double <- function(x, to, ...) percent(x)  # use helper
vec_cast.double.vctrs_percent <- function(x, to, ...) vec_data(x) # strip the attributes 

# check it works
vec_cast(0.5, percent())
vec_cast(percent(0.5), double())
```

## `vec_c` `[<-` and `[[<-`

```{r}
vec_c(percent(0.5), 1)
vec_c(NA, percent(0.5))
# but
vec_c(TRUE, percent(0.5))

x <- percent(c(0.5, 1, 2))
x[1:2] <- 2:1
x[[3]] <- 0.5
x
```

For convenience, define

```{r}
as_percent <- function(x) {
  vec_cast(x, new_percent())
}
```

## Additional methods

Occasionally, it is useful to provide conversions that go beyond what’s allowed in casting.In this case, `as_percent()` should be generic, the default method should cast, and then additional methods should implement more flexible conversion:

```{r}
as_percent <- function(x, ...) {
  UseMethod("as_percent")
}

as_percent.default <- function(x, ...) {
  vec_cast(x, new_percent())
}

as_percent.character <- function(x) {
  value <- as.numeric(gsub(" *% *$", "", x)) / 100
  new_percent(value)
}
```

# Building S3 Methods with {vctrs}

-   Defining S3 classes and constructors

-   Using `UseMethod()` for method dispatch

-   Handling inheritance and `NextMethod()`

-   The role of default methods

-   Preserving subclass information with {vctrs}

## 1. Defining the secret S3 Class

-   We start by defining our **base class** secret.\
    It stores a numeric vector but hides the actual numbers when printed.

-   As a first step, define a low-level **constructor**. The vctrs package provides `new_vctr()` to simplify the creation of a custom vector class.

```{r}
library(vctrs)

new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

# Helper function to create a secret vector
secret <- function(x) {
  if (!is.double(x)) stop("`x` must be a numeric (double) vector")
  new_secret(x)
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))  # Masks the numbers when printed
  invisible(x)
}

# Example usage
x <- secret(c(15, 1, 456))
print(x)
class(x)

# TODO: ADD VALIDATOR, ADD MORE ATTRBUTES TO MAKE USE OF VCTRS NEW_VCTR

```

## 2. Creating an S3 Generic Function with `UseMethod()`

let's create a generic function that will "reveal" the numbers hidden by the "x" representation.\
`UseMethod("reveal")` ensures that reveal.secret() is used when x is a "secret".

```{r}

# Define a generic function for revealing secrets
reveal <- function(x) {
  UseMethod("reveal")
}

# Method for objects of class 'secret'
reveal.secret <- function(x) {
  as.numeric(x)  # Reveal the original numeric values
}

# Example usage
x <- secret(c(15, 1, 456))
reveal(x)  # Output: 15 1 456


```

## Default method for other classes

If x is not a "secret", `reveal.default()` is used instead.

```{r}
# Default method (fallback for other classes)
reveal.default <- function(x) {
  x  # Return as is
}

y <- c(100, 200)
reveal(y)  # Uses reveal.default(), returns unchanged
```

## 3. Inheritance and NextMethod()

When we define a subclass like `supersecret`, it **inherits** all properties and behaviors of its parent class secret. However, it can also introduce **additional attributes or behaviors**.

**Why use NextMethod()?** \* It ensures that the superclass method is properly called. \* It allows supersecret to inherit behavior from secret.

```{r}
new_supersecret <- function(x = double(), level = "high") {
  stopifnot(is.double(x))
  structure(x, class = c("supersecret", "secret"), level = level)
}

y <- new_supersecret(c(99, 1000))

attributes(y)

# todo: modify to use constructor of original class
```

Now, `supersecret` has an extra attribute (level).

## How method dispatch work in this case

Let’s define a method for printing `supersecret` objects:

```{r}
print.supersecret <- function(x, ...) {
  cat("Super Secret! Hidden level:", attr(x, "level"), "\n")
  NextMethod()  # Calls the parent print method (print.secret)
}

y <- new_supersecret(c(99, 1000), level = "top")
print(y)
class(y)

# TODO: show method dispatch
```

First, `print.supersecret()` is called (specific to `supersecret`).\
Then, `NextMethod()` ensures that `print.secret()` still runs.

## Additional methods

Let's define an S3 generic function that behaves differently based on the class.\
Example: Masking Numbers More for `supersecret` objects

```{r}
mask_number <- function(x) {
  UseMethod("mask_number")
}

mask_number.secret <- function(x) {
  strrep("x", nchar(x))
}

mask_number.supersecret <- function(x) {
  strrep("*", nchar(x))  # Stronger masking for supersecret
}

mask_number(s)   # "xx"  "x"   "xxx"
mask_number(ss)  # "**"  "*"   "***"


```

If an object doesn’t match any specific method, R falls back to **default methods**.

```{r}
mask_number.default <- function(x) {
  stop("mask_number() is not implemented for this object type.")
}
mask_number("hello")
# Error: mask_number() is not implemented for this object type.

```

This prevents unexpected errors for unsupported types.

## 4. Subsetting and Keeping the Subclass

When we define a subclass (`supersecret`) of secret, it inherits all methods of secret. However, if we subset the object, the subclass information gets lost.

Let's consider the current situation:

```{r}
x <- new_supersecret(c(15, 1, 456))
class(x)

x_subset <- x[1:2]
class(x_subset)

```

**Why does this happen?**

-   By default, the subsetting operator `([)` from base R drops attributes

-   The class vector is an attribute, so unless we manually reapply it, subsetting a supersecret object returns just a secret object

## Using `vctrs::vec_restore()`

Instead of manually setting the class, we can use:

```{r}

`[.secret` <- function(x, ...) {
  out <- NextMethod()  # Perform subsetting
  vctrs::vec_restore(out, x)  # Restore subclass information
}

```

It automatically restores all attributes

```{r}
x_subset <- x[1:2]
class(x_subset)

attributes(x_subset)
```

## 5. Handling concatenation

When combining different objects, we need to ensure type consistency. When working with custom S3 classes, you might need to combine objects (e.g., using c() in R). However, standard concatenation strips the class attributes, which can break subclass behavior.

```{r}
x <- new_secret(c(1, 2))
y <- new_supersecret(c(3, 4))

c(x, y)  
# [1] 1 2 3 4  (Lost class attribute! Just a numeric vector now)

```

You can **define type behavior** with `vec_ptype2()`. basically, define rules for mixing `secret` and `supersecret`

```{r}
vec_ptype2.secret.secret <- function(x, y, ...) new_secret()
vec_ptype2.supersecret.secret <- function(x, y, ...) new_supersecret()
vec_ptype2.secret.supersecret <- function(x, y, ...) new_supersecret()

vec_c.secret <- function(...) {
  new_secret(vec_c(!!!list(...)))
}

vec_ptype2(new_secret(), new_secret())        
# Returns a new_secret object ✅

vec_ptype2(new_supersecret(), new_secret())  
# Returns a new_supersecret object ✅

```

## Implement Safe Concatenation with `vec_c()`
Now that we’ve defined type rules, we need to tell {vctrs} how to concatenate objects while preserving class attributes.

```{r}
vec_c.secret <- function(...) {
  new_secret(vec_c(!!!list(...)))
}

vec_c.supersecret <- function(...) {
  new_supersecret(vec_c(!!!list(...)))
}

c(x, y)  
vec_c(x, y)
```

## What if we combine numeric and secret?
We can also define what happens if a base numeric value is concatenated with a secret object
```{r}
vec_ptype2.secret.double <- function(x, y, ...) new_secret()
vec_ptype2.double.secret <- function(x, y, ...) new_secret()

```

This ensures that even if a raw number is added, the output remains a secret!
```{r}
vec_c(new_secret(1:3), 4)
# Returns new_secret object ✅
```

## 6. Handling arithmetic operations 

We can define how these operations should behave for our custom objects. This is done using the `vec_arith` method provided by the {vctrs} package. For example, if you want to add two secret objects, you need to tell R how to perform the addition.

```{r}
# Defining vec_arith() for secret and supersecret
vec_arith.secret <- function(op, x, y, ...) {
  # Ensure y is numeric
  if (!is.numeric(y)) stop("Arithmetic is only supported with numeric types")
  
  # Perform the arithmetic using base function and wrap it in `new_secret`
  new_secret(vec_arith_base(op, x, y))
}

vec_arith.supersecret <- function(op, x, y, ...) {
  # Ensure y is numeric
  if (!is.numeric(y)) stop("Arithmetic is only supported with numeric types")
  
  # Perform the arithmetic using base function and wrap it in `new_supersecret`
  new_supersecret(vec_arith_base(op, x, y))
}

# Example
x <- new_secret(c(5, 10, 15))
y <- new_secret(c(1, 2, 3))

result <- x + y  # This triggers `vec_arith.secret`
print(result)

x <- new_secret(c(5, 10, 15))
y <- new_supersecret(c(1, 2, 3))

result <- x + y  # This triggers `vec_arith.supersecret`
print(result)

```

