---
title: "S3 Methods in R - Part 1"
author: Rossana Tatulli
format: revealjs
revealjs:
  slideNumber: true
  theme: simple
  transition: fade
  fontSize: 20
  slideWidth: 1280
  slideHeight: 720
editor: visual
echo: TRUE
---

## Intro on OOP and S3 methods

-   **OOP**: based on objects, which have a class, - *what the object is* -, and a method - *what the object does*
-   classes are organized into **hierarchies** so that if a method is missing for a certain class, it is inherited from the parent
-   Main goal of OOP: **polymorphism**, so that you can consider the function interface separately from the function implementation
-   S3 methods are most commonly used OOP system in CRAN packages, they are an implementation of ***functional OOP***:
    -   methods belong to generic functions
    -   method calls look like ordinary functions calls

## The basics

-   An S3 object is a base type with at least a `class` attribute(e.g., factor)

-   It behaves differently from its underlying base type whenever it’s passed to a **generic function**, which**:**

    -   uses a different implementation (the **method**) depending on the class of an argument

    -   finds that method by performing **method dispatch**

You can use [`sloop::s3_dispatch()`](https://sloop.r-lib.org/reference/s3_dispatch.html) to see the process of method dispatch:

```{r echo = TRUE}
library(sloop)
f <- factor(c("a", "b", "c"))
s3_dispatch(print(f))
```

## Classes

-   no formal definition of a class, to make an object an instance of a class, you simply set the **class attribute**

```{r}
# Create and assign class in one step
x <- structure(list(), class = "my_class")

# Create, then set class
x <- list()
class(x) <- "my_class"
```

-   You can determine the class of an S3 object with [`class(x)`](https://rdrr.io/r/base/class.html), and see if an object is an instance of a class using [`inherits(x, "classname")`](https://rdrr.io/r/base/class.html).

```{r}
class(x)
inherits(x, "my_class")

inherits(x, "your_class")
```

## **Class name guidelines**: {.smaller}

-   **Any string can be a class name**, but it’s advised to keep it simple by using only **letters** and **underscores** (`_`). Avoid using periods (`.`) in the class name because it could be confused with the period separator used between a generic name and a class name in R (e.g., `generic.class`).

-   When using classes within a package, it’s a good idea to **include the package name in the class name** (e.g., `mypackage_myclass`). This reduces the risk of class name collisions

## **Three recommended functions for class management**: {.smaller}

1\. **Constructor: `new_myclass()`**

-   **Purpose**: The constructor is a **low-level function** that is responsible for **creating new objects** of your class with the correct internal structure.

2\. **Validator: `validate_myclass()`**

-   **Purpose**: The validator is used to **check** if the object’s internal values are valid or meet certain conditions.

-   **What it does**: It performs more **computationally expensive checks** to ensure that the object's fields contain **correct data**

3\. **Helper: `myclass()`**

-   **Purpose**: The helper function is a **convenience function** for users to easily create objects of your class. It is typically a wrapper around the constructor function.

## TODO: add examples on contructor validator and helper

## Generics and methods - method dispatch {.smaller}

-   every generic calls `UseMethod` to perform method dispatch. `UseMethod()` takes two arguments: the name of the generic function (required), and the argument to use for method dispatch (optional)
-   it creates a vector of method names, `paste0("generic", ".", c(class(x), "default"))`, and then looks for each potential method in turn..see this in action with `sloop::s3_dispatch()`
-   “default” class is not a real class, but is included to define a standard fallback that is found whenever a class-specific method is not available.

## Finding methods {.smaller}

To find all methods defined for a generic or associated with a class, use `sloop::s3_methods_generic()` and `sloop::s3_methods_class()`

ADD examples

## Creating methods {.smaller}

-   **Best Practice**: Only create methods for generics or classes that you own (or collaborate with the author of the package/class to add your method).
-   **Argument Matching:** Your method should have the same argument signature as the generic, except when using ... for flexibility.
-   **Ellipsis (...)**: Using ... in methods allows you to accept additional arguments, but be careful because misspelled arguments will be silently ignored, which could cause hidden issues.

## TODO: add examples, maybe the print i developed for syncdr

------------------------------------------------------------------------

Exercises to add

------------------------------------------------------------------------

## Object styles

Think about if relevant

## Inheritance {.smaller}

Mechanism through which **different classes** can **share behavior**, powered by 3 ideas:

1\. Class can be a character vector

2\. If method not found for the first class, R looks to method for the second class

3\. A method can delegate work by calling `NextMethod()`, "-\>"

---

```{r}

# Example with ordered 
# ordered() creates an ordered factor with a given input 

# 1. ----- #
ordered(1:4)
class(ordered(1:4))

# (Even though both objects contain the same values, ordered() ensures that comparisons (<, >) are meaningful:)

# 2. ----- #
x <- ordered(1:4)
s3_dispatch(print(ordered("x")))

# 3. ----- #
s3_dispatch(ordered("x")[1])

```

------------------------------------------------------------------------

***Output takeaways***

[*Step 1: Printing*]{.underline}

\- The print() function is called on the result of ordered("x")\[1\].

\- R first tries print.factor, which is the specialized print method for factors (including ordered factors).

\- The =\> symbol means this method was used.

\- The \* before `print.default` means it was skipped, because `print.factor` was found.

--\> Since an ordered factor is also a factor, R uses print.factor to display it

------------------------------------------------------------------------

[*Step 2: Subsetting*]{.underline}

\- When `ordered("x")[1]` is evaluated, R looks for a `[.ordered` method (a special method for subsetting ordered factors).

\- No =\> appears next to `[.ordered`, meaning it was not found.

\- R then falls back to `[.factor`, which exists and is used (=\> means this method was applied)

\- `-> [ (internal)` means that the base "'['" operator was used at the lowest level

--\> Ordered factors don’t have a specialized `[.ordered` method, so they rely on `[.factor`.

--\> The actual subsetting operation is ultimately handled by the internal `[` function.

------------------------------------------------------------------------

::: {.callout-caution collapse="true"}
## Why delegation to `-> [ (internal)`

In R’s S3 method dispatch system:

-   R first looks for `[.ordered`, but it doesn't exist.
-   It then finds `[.factor` and uses it (=\> means it was applied).
-   Inside `[.factor`, R calls `NextMethod()`, which means:
    -   It delegates the call to `[.default` (if it exists) or the base `[` function.
    -   Since `[.default` is not used here, it goes directly to `[ (internal)`.

--\> 💡 Key takeaway:

`-> [ (internal)` means that `[.factor` did not completely handle the operation itself but instead delegated the actual work to the base R \``[` function (written in C). This delegation is likely happening via `NextMethod()`
:::

## Relationship between the classes that appear together in a class vector {.smaller}

In our example, `ordered` is a **subclass** of `factor` and `factor`is a **superclass** of `ordered`.

It is recommend that you adhere to two principles when creating a subclass:

1.  Sub-class and class belong to the **same base type**
2.  **Attributes** of the subclass should be a **superset** of the attributes of the **superclass**.

---

**Example**

```{r}
# Function to create a "labelled_factor" subclass
labelled_factor <- function(x, label) {
  # Ensure the base type remains the same as the superclass (factor)
  if (!inherits(x, "factor")) {
    stop("x must be a factor")
  }
  
  # Add an extra attribute ("label") while preserving existing attributes
  structure(x, class = c("labelled_factor", "factor"), label = label)
}

# Example usage
x  <- factor(c("low", "medium", "high"))
lf <- labelled_factor(x, label = "Risk Level")

# Check structure
str(lf)
```

## `NextMethod()` {.smaller}

Let's start with an example

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

---

However, the default `[` method does not preserve the class

```{r}
s3_dispatch(x[1])
x[1]

```

---

Need to define a `[.secret` method:

Option 1:

```{r}
`[.secret` <- function(x, i) {
  new_secret(x[i])
}
```

Option 2:

```{r}
`[.secret` <- function(x, i) {
  x <- unclass(x)
  new_secret(x[i])
}
x[1]
```

::: {.callout-note}

-   Can you guess why [Option 1]{.underline} does not work?

-   Why is [Option 2]{.underline} not efficient? (hint: *copy-on-modify* mechanism)

:::

---

With `NextMethod()` instead

```{r}
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}
x[1]
s3_dispatch(x[1])
```

-   **`NextMethod()` tracks the list of potential next methods:** this means that R keeps track of the method dispatch chain, which consists of all the possible methods that could be invoked next for the object being processed

-   When a method is called, R checks the class of the object and finds the corresponding method. If the method defines a call to `NextMethod()`, R will **continue searching for the next method in the chain**, considering the inheritance hierarchy and class structure

## Allowing sub-classes {.smaller}
 
**Subclassing** might allows for more flexible and extensible code. To allow subclasses, the parent constructor needs to have`...`and `class` arguments:

```{r}
new_secret <- function(x, ..., class = character()) {
  stopifnot(is.double(x))

  structure(
    x,
    ...,
    class = c(class, "secret")
  )
}
#note that the class argument is used to attach the class to the object, and "secret" is added to the class chain.
```

---

-   Subclasses (like `supersecret`) can be created by simply calling the parent constructor with the `class` argument modified to the subclass name.

```{r}
new_supersecret <- function(x) {
  new_secret(x, class = "supersecret")
}

```

---

To allow inheritance, you need to think about your methods, as you can no longer use the constructor. See the example below:

```{r}
x2 <- new_supersecret(c(15, 1, 456))
x2

`[.secret` <- function(x, ...) {
  new_secret(NextMethod())
}

x2[1:3]
```

We need to refine the `[.secret` method, because currently it always return a `secret()` even when given a supersecret

-   need to use the vctrs package –\> we will see thisin the next session

-   general idea: using `vctrs::vec_restore()` helps manage subclassing: when an object loses its subclass information (e.g., after subsetting), it can be restored to the correct class.

## More on `UseMethod`, `NextMethod`, and `default` method in S3 {.smaller}

`UseMethod()`: 

* the mechanism that triggers method dispatch in S3 -dispatch appropriate function based on the class of the object passed to it
* It's used in the **generic function**

---

`NextMethod()`:

* used inside a method to call the next method in the dispatch chain
* If a method is called for a subclass, the `NextMethod()` function is used to call the corresponding method of the superclass, keeping the method dispatch chain intact

---

`default` methods:

* Fallback methods that are used when no more specific method is found for a particular class
* **useful** when you want to provide a generic implementation of a function for all object types, while still allowing specific classes to override the method with specialized behavior
* **necessary** if you want to ensure that a function works for any object, even if that object doesn’t have a specific method defined for it
* **unnecessary** when you want to ensure specific behavior for each subclass or class in a way that does not fall back on a generic implementation

--- 

Example:

```{r define, error=TRUE}
# Define generic class
area <- function(x) {
  UseMethod("area", x)
}

# Define the specific methods
# Circle area
area.circle <- function(x) {
  pi * x$radius^2
}
# Square area
area.square <- function(x) {
  x$side^2
}
# Rectangle area
area.rectangle <- function(x) {
  x$length * x$width
}

# Define the Default Method
area.default <- function(x) {
  stop("No method for calculating area for this class.")
}
```

---
```{r call-objs, error=TRUE}
# Create some objects and call the function

# Create a circle
circle <- list(radius = 5)
class(circle) <- "circle"

# Create a square
square <- list(side = 4)
class(square) <- "square"

# Create a rectangle
rectangle <- list(length = 6, width = 3)
class(rectangle) <- "rectangle"

# Create a generic object (no specific class)
unknown_shape <- list()
class(unknown_shape) <- "unknown"

# ________ ####
# Calling area for specific shapes
cat("Area of circle:", area(circle), "\n")        # Should calculate area of the circle
cat("Area of square:", area(square), "\n")        # Should calculate area of the square
cat("Area of rectangle:", area(rectangle), "\n")  # Should calculate area of the rectangle

# Calling area for an object with no specific class
cat("Area of unknown shape:", area(unknown_shape), "\n")  # Should trigger default method

```



## S3 methods - application with {vctrs}

NEXT SESSION
