---
title: "s3_methods"
format: 
  revealjs:
    center: true
editor: visual
echo: TRUE
---

## Intro on OOP and S3 methods

-   **OOP**: based on objects, which have a class, - *what the object is* -, and a method - *what the object does*
-   classes are organized into **hierarchies** so that if a method is missing for a certain class, it is inherited from the parent
-   Main goal of OOP: **polymorphism**, so that you can consider the function interface separately from the function implementation
-   S3 methods are most commonly used OOP system in CRAN packages, they are an implementation of ***functional OOP***:
    -   methods belong to generic functions
    -   method calls look like ordinary functions calls

## The basics

-   An S3 object is a base type with at least a `class` attribute(e.g., factor)

-   It behaves differently from its underlying base type whenever it’s passed to a **generic function**, which**:**

    -   uses a different implementation (the **method**) depending on the class of an argument

    -   finds that method by performing **method dispatch**

You can use [`sloop::s3_dispatch()`](https://sloop.r-lib.org/reference/s3_dispatch.html) to see the process of method dispatch:

```{r echo = TRUE}
library(sloop)
f <- factor(c("a", "b", "c"))
s3_dispatch(print(f))
```

## Classes

-   no formal definition of a class, to make an object an instance of a class, you simply set the **class attribute**

```{r}
# Create and assign class in one step
x <- structure(list(), class = "my_class")

# Create, then set class
x <- list()
class(x) <- "my_class"
```

-   You can determine the class of an S3 object with [`class(x)`](https://rdrr.io/r/base/class.html), and see if an object is an instance of a class using [`inherits(x, "classname")`](https://rdrr.io/r/base/class.html).

```{r}
class(x)
inherits(x, "my_class")

inherits(x, "your_class")
```

## **Class name guidelines**: {.smaller}

-   **Any string can be a class name**, but it’s advised to keep it simple by using only **letters** and **underscores** (`_`). Avoid using periods (`.`) in the class name because it could be confused with the period separator used between a generic name and a class name in R (e.g., `generic.class`).

-   When using classes within a package, it’s a good idea to **include the package name in the class name** (e.g., `mypackage_myclass`). This reduces the risk of class name collisions

## **Three recommended functions for class management**: {.smaller}

1\. **Constructor: `new_myclass()`**

-   **Purpose**: The constructor is a **low-level function** that is responsible for **creating new objects** of your class with the correct internal structure.

2\. **Validator: `validate_myclass()`**

-   **Purpose**: The validator is used to **check** if the object’s internal values are valid or meet certain conditions.

-   **What it does**: It performs more **computationally expensive checks** to ensure that the object's fields contain **correct data**

3\. **Helper: `myclass()`**

-   **Purpose**: The helper function is a **convenience function** for users to easily create objects of your class. It is typically a wrapper around the constructor function.

## TODO: add examples on contructor validator and helper

## Generics and methods - method dispatch

-   every generic calls `UseMethod` to perform method dispatch. `UseMethod()` takes two arguments: the name of the generic function (required), and the argument to use for method dispatch (optional)
-   it creates a vector of method names, `paste0("generic", ".", c(class(x), "default"))`, and then looks for each potential method in turn..see this in action with `sloop::s3_dispatch()`
-   “default” class is not a real class, but is included to define a standard fallback that is found whenever a class-specific method is not available.

## Finding methods

To find all methods defined for a generic or associated with a class, use `sloop::s3_methods_generic()` and `sloop::s3_methods_class()`

ADD examples

## Creating methods

-   **Best Practice**: Only create methods for generics or classes that you own (or collaborate with the author of the package/class to add your method).
-   **Argument Matching:** Your method should have the same argument signature as the generic, except when using ... for flexibility.
-   **Ellipsis (...)**: Using ... in methods allows you to accept additional arguments, but be careful because misspelled arguments will be silently ignored, which could cause hidden issues.

## TODO: add examples, maybe the print i developed for syncdr

------------------------------------------------------------------------

Exercises to add

------------------------------------------------------------------------

## Object styles

Think about if relevant

## Inheritance

Mechanism through which **different classes** can **share behavior**, powered by 3 ideas:

1\. Class can be a character vector

2\. If method not found for the first class, R looks to method for the second class

3\. A method can delegate work by calling `NextMethod()`, "-\>"

---

```{r}

# Example with ordered 
# ordered() creates an ordered factor with a given input 

# 1. ----- #
ordered(1:4)
class(ordered(1:4))

# (Even though both objects contain the same values, ordered() ensures that comparisons (<, >) are meaningful:)

# 2. ----- #
x <- ordered(1:4)
s3_dispatch(print(ordered("x")))

# 3. ----- #
s3_dispatch(ordered("x")[1])

```

------------------------------------------------------------------------

***Output takeaways***

*Step 1: Printing*

\- The print() function is called on the result of ordered("x")\[1\].

\- R first tries print.factor, which is the specialized print method for factors (including ordered factors).

\- The =\> symbol means this method was used.

\- The \* before `print.default` means it was skipped, because `print.factor` was found.

--\> Since an ordered factor is also a factor, R uses print.factor to display it

------------------------------------------------------------------------

*Step 2: Subsetting*

\- When `ordered("x")[1]` is evaluated, R looks for a `[.ordered` method (a special method for subsetting ordered factors).

\- No =\> appears next to `[.ordered`, meaning it was not found.

\- R then falls back to `[.factor`, which exists and is used (=\> means this method was applied)

\- `-> [ (internal)` means that the base \`\[ operator was used at the lowest level

--\> Ordered factors don’t have a specialized \[.ordered method, so they rely on \[.factor.

--\> The actual subsetting operation is ultimately handled by the internal \`\[ function.

------------------------------------------------------------------------

::: {.callout-caution collapse="true"}
## Why delegation to `-> [ (internal)`

In R’s S3 method dispatch system:

-   R first looks for \[.ordered, but it doesn't exist.
-   It then finds \[.factor and uses it (=\> means it was applied).
-   Inside \[.factor, R calls NextMethod(), which means:
    -   It delegates the call to \[.default (if it exists) or the base \`\[ function.
    -   Since \[.default is not used here, it goes directly to \[ (internal).

--\> 💡 Key takeaway:

`-> [ (internal)` means that `[.factor` did not completely handle the operation itself but instead delegated the actual work to the base R \``[` function (written in C). This delegation is likely happening via `NextMethod()`
:::

## S3 methods - application with {vctrs}

NEXT SESSION
